### 2022-09-19 chapters/optimization/python-dir/python.tex
Proofreading

In particular,\citeauthor{noureddine_preliminary_2012}~\cite{noureddine_preliminary_2012} in 2012, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
According to their experiments, an interpreted programming language, like Python, can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. \Cref{fig:hannoi}).
In this chapter, we explore the oblivious optimizations that can be applied to Python legacy applications to reduce their energy footprint.
As Python is widely adopted by software services deployed in public and private cloud infrastructures, we believe that our contributions will benefit a wide diversity of legacy systems and not only favorably contribute to reducing the carbon emissions of ICT, but also reduce their cloud invoice for the resources consumed by these services.
More specifically, this chapter focuses on runtime optimizations that can be adopted by developers to leverage the power consumption of Python applications.
We start by studying the impact of programmer choices, such as the type of data or control structures, on the global energy consumption of the execution code.
Then, we discuss other factors, such as the levels of concurrency, before we investigate other non-intrusive approaches to optimize the energy consumption of applications.
One type of optimization includes alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.

@@

\subsection{Python Popularity}
Nowadays, Python seems to attract a large community of developers who are interested in data analysis, web development, system administration, and machine learning.
According to a survey conducted in 2018 by JetBrains,\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}} one can fear that the wide adoption of dynamic programming languages, like Python, in production may critically hamper the power consumption of ICT.
As the popularity of such dynamic programming languages partly builds on the wealth and the diversity of their ecosystem (\emph{e.g.}, the NumPY, SciKit\,Learn, and Panda libraries in Python), one cannot reasonably expect that developers will likely move to an alternative programming language mostly for energy considerations.
Rather, we believe that a better option consists of leveraging the strength of this rich ecosystem to promote energy-efficient solutions to improve the power consumption of legacy software systems.

\subsection{Python Gluttony}
% In the other hand,because of the interpreted nature of Python code, this programming language has paid a high price in terms of performance.  \ref{fig:clbg}, memory and energy consumption.
According to~\cite{pinto_energy_2017} and~\cite{noureddine_preliminary_2012}, Python tends to be one more energy hungry programming language.
As one can notice in \Cref{fig:hannoi}, Python consumes $30$ times more than C or C++.
The benchmark was done with an implementation of the \fnurl{Tower of Hanoi}{ https://en.wikipedia.org/wiki/Tower_of_Hanoi} of 30 disks.

Python consumes a lot of energy, mainly because it is slow in execution.
Its flexibility and simplicity caused it to drop off in performance because Python gains its flexibility from being a dynamic language.
Therefore, it needs an interpreter to execute its programs, which makes them much slower compared to the others that are written in compiled programming languages, such as C and C++ or semi-compiled languages like Java.

@@

@@

\subsection{Python Limits}
To reduce the energy consumption of Python, we started by targeting the main usage of this programming language, which is revealed to be data science and web development.
\Cref{fig:usecase} illustrates a study published by the JetBrain company on Python developers\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}.
57\% of the respondents reported that they use Python for data science, 51\% said they are using it for web development, and around 40\% are using it for system administration.
\paragraph{Remark}
The options in this survey were not mutually exclusive. As a result, the total of the percentages is greater than $100\%$.

@@

In this chapter, we investigate the energy footprint of Python in its most popular domains of adoption.
We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{hasan_energy_2016} did in~\cite{hasan_energy_2016}.
Then, we measure the energy consumption of several Python implementations to propose a non-intrusive technique to improve energy efficiency.
Therefore, this chapter will address the following research questions:
\begin{compactenum}[\indent\bf RQ\,1:]
    \item \emph{What is the energy footprint of Python when used in data science?}
    \item \emph{Are the Python guidelines energy efficient by construction?}
    \item \emph{Can we reduce the energy consumption of Python programs without altering the source code?}
\end{compactenum}

@@

Django~\footnote{\url{https://www.djangoproject.com/}} and Flask\footnote{\url{https://flask.palletsprojects.com/}}  are the most popular frameworks for web development. According to a Jetbrains poll\footnote{\url{https://lp.jetbrains.com/python-developers-survey-2021/}} , Django is used in 40\% of the cases, while Flask is used in 41\% of the cases.

In contrast to Flask, which is still a microframework, Django is a high-level web framework that provides a standard method for creating and maintaining complex and scalable database-driven websites quickly and effectively. Therefore, this study will focus on The latter one.

\subsection{life cycle of a request in Django}

Django is a MVT (Model-View-Template) framework, which means that it follows the MVC (Model-View-Controller) pattern. Figure~\ref{fig:django-life-cycle} describes the life cycle of a request in this framework.  Whenever a request arrives in Django, it is processed by several middlewares, one at a time. These middlewares are in charge of authentication, security, and so on.
Once the request is processed by these middlewares, it is passed to the URL Router, which will simply extract the URL from the request and will try to match it to the defined URLs.
After getting the matching URL, the corresponding view function is called. This function is responsible for treating the request, gathering the data, and then generating the response that will be put inside a template to be returned as a response.
However, as mentioned before, Django is an MVT model. which offers an automatic way to retrieve data from the database to the view using the \emph{Object Relational Mapping} (ORM)~\cite{o2008object}.

@@

First, we start by investigating the energy consumption of the request-response life cycle in Django, in ordrer to determine which part consumes the most energy. To do so, we created a simple Django application that simply returns the response of the request. and then tracked its energy consumption using joulehunter~\footnote{\url{https://github.com/powerapi-ng/joulehunter}}.

Joulehunter is an open source library that we developed in order to help practitioners find the energy hotspots in their applications using statistical profilers. In the case of Django, it can be added as middleware without any extra setup or changes to the source code. The energy consumption of the request-response life cycle in Django is shown in the figure~\ref{fig:django_life_cycle_naive}.As we can notice, 91.4\% of the total energy consumption is spent to resolve the request by retrieving the data while only 5.27\% is spent to render the response.

@@

%% REMARQUE the experiment with joulehunter isn't conducted within the same machine that is used for generating the boxplot graphs 
% We have chosen Django as our web framework.
Therefore,we chose to observe if the choice of the database and the ORM impacts the energy consumption and the performance of the website.
\subsection{Experiment \& Results}
To accomplish so, we examined the cost of a single request of the prior website using various ways to extract data from the database.
We considered using two different databases, \textsc{PostgreSQL} and \textsc{SQlite3}, that store the same records, and three different ways to fetch the records.
\begin{enumerate}
    \item \textsf{Vanilla} relies only on the ORM to retrieve the data,
    \item \textsf{Prefetch} queries the data before being requested,
    \item \textsf{Optimized} leverages SQL without passing by the ORM.
\end{enumerate}

%%%%% have to rephrase this 
As one can observe in \Cref{fig:django}, the strategy to query stored data has a huge impact on the energy consumption.
As the \textsf{Vanilla} strategy can consume up to $10\times$ more energy than the \textsf{Optimized} one.
Conversely, the choice of the database does not exhibit a key impact on the total energy despite their different behavior regarding the execution time and the average power.
This can be useful to support developers in choosing which database engine they can adopt, based on the number of expected requests and the targeted performance.

@@

Another interesting observation is the impact of the interpreter, as \Cref{fig:django} highlights.
For example, using the \textsf{Pypy} interpreter reduces the energy consumption, even when we adopting the \textsf{Vanilla} strategy.

Finally, we run the same experiment with the \textsf{Optimized} strategy
using joulehunter
.Figure~\ref{fig:django_profiled_optimized} shows the new graph of the energy consumption. As one can notice. while the rendering method consumed the same amount of energy as in  the \textsf{Vanilla} strategy (arround 1.3KJ), the resolve part droped by $20\times$.

\subsection{Conclusion}
We conclude this study by stating that the database and ORM selection have a major impact on the website's performance and energy consumption.
And unlike the rendering portion, there are many options available for the database, the ORM, and the data handling strategy.
\begin{figure}[hbt]
    \centering
    \includegraphics[width=\linewidth]{imgs/django_profiled_optimized}
    \caption{Tree representation of the energy consumption of a single request in django (naive version)}
    \label{fig:django_profiled_optimized}
\end{figure}

\Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications when introducing parallelism.
Python applications are single-threaded systems constrained by the \emph{global lock system} (GLS).
However, due to the increase of cores/threads per CPU, many Python libraries started to take advantage of this hardware feature by allowing concurrent execution of Python processes.
For example, the multiprocessing library\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}} spawns sub-processes to increase the degree of concurrency of Python applications.
As one can see in \Cref{fig:python_multiprocessing}, the energy consumption is correlated with the number of threads until reaching the limit of physical cores, when concurrent processes start to compete for the CPU.

@@

\section{Python Insights}
This section aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
Therefore, we start by extending the work of \citeauthor{hasan_energy_2016} and
\citeauthor{oliveira_recommending_nodate} to the context of Python environments
\cite{hasan_energy_2016,oliveira_recommending_nodate}.
\subsection{experiment and results}
%%%% concurency 
Another field of investigation is the type of data and control structures that might impact the energy consumption.
We iterate over a list using 3 methods.
First, the classical \texttt{for(i in range(len(n)))}.
However, as we can see here, unlike other programming languages, it requires extra operations, such as determining the length of the collection and then using the iterator range.
So, we tried the more adapted version \texttt{for(element in collection)}.
Moreover, in most programming languages. the \texttt{for} loop is translated to a \texttt{while} loop ( transformation from D type to B type -- asm -- ), therefore we wanted to compare this with a \texttt{while} version.
After determining the main ways to iterate over a \texttt{loop}, we run the collection algorithms of different data types to observe if they impact the energy consumption of the code, and we repeated it for the size of the collection.

@@

\subsection{Conclusion}
This study demonstrates that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.

\section{Python \& Multiprocessing}
\Cref{fig:python_multiprocessing} compares the behavior of python programs when we try to introduce parallelism.
As we know python is a single-threaded program thanks to the GLS (\emph{Global Lock System}) however, due to the increase in the number of cores/threads per CPU, many libraries started to take advantage of this hardware feature by simulating multithreading as multiple instances of the processor. % TODO: TO VERIFY

>>>>

In particular, \citeauthor{noureddine_preliminary_2012} in 2012~\cite{noureddine_preliminary_2012}, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
According to their experiments, an interpreted programming language, like Python, can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. \Cref{fig:hannoi}).
In this chapter, we explore the oblivious optimizations that can be applied to Python legacy applications to reduce their energy footprint.
As Python is widely adopted by software services deployed in public and private cloud infrastructures, we believe that our contributions will benefit a wide diversity of legacy systems and not only favorably contribute to reducing the carbon emissions of ICT, but also reduce their cloud invoice for the resources consumed by these services.
More specifically, this chapter focuses on runtime optimizations that can be adopted by developers to leverage the power consumption of Python applications.
We start by studying the impact of programmer choices, such as the type of data or control structures, on the global energy consumption of the execution code.
Then, we discuss other factors, such as the levels of concurrency, before we investigate other non-intrusive approaches to optimize the energy consumption of applications.
One type of optimization includes alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.

@@

\subsection{Python Popularity}
Nowadays, Python attracts a large community of developers who are interested in data analysis, web development, system administration, and machine learning.
According to a survey conducted in 2018 by JetBrains,\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}} one can fear that the wide adoption of dynamic programming languages in production, like Python, may critically hamper the power consumption of ICT.
As the popularity of such dynamic programming languages partly builds on the wealth and the diversity of their ecosystem (\emph{e.g.}, the NumPY, SciKit\,Learn, and Panda libraries in Python), one cannot reasonably expect that developers will likely move to an alternative programming language mostly for energy considerations.
Rather, we believe that a better option consists of leveraging the strength of this rich ecosystem to promote energy-efficient solutions to improve the power consumption of legacy software systems.

\subsection{Python Gluttony}
% In the other hand,because of the interpreted nature of Python code, this programming language has paid a high price in terms of performance.  \ref{fig:clbg}, memory and energy consumption.
According to~\cite{pinto_energy_2017} and~\cite{noureddine_preliminary_2012}, Python tends to be one more energy hungry programming language.
As one can notice in \Cref{fig:hannoi}, Python consumes $30$ times more than C or C++.
The benchmark was done with an implementation of the \fnurl{Tower of Hanoi}{https://en.wikipedia.org/wiki/Tower_of_Hanoi} of 30 disks.

@@

Python consumes energy, mainly because it is slow in execution.
Its flexibility and simplicity caused it to drop off in performance because Python gains its flexibility from being a dynamic language.
Therefore, it requires a faster interpreter to execute its programs to compete against alternatives written in compiled programming languages, such as C and C++ or semi-compiled languages like Java.

@@

\subsection{Python Limits}
To reduce the energy consumption of Python, we started by targeting the main usage of this programming language, which is revealed to be data science and web development.
\Cref{fig:usecase} illustrates a study published by the JetBrain company on Python developers.\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}
57\% of the respondents reported that they use Python for data science, 51\% said they are using it for web development, and around 40\% are using it for system administration.\footnote{The options in this survey were not mutually exclusive.As a result, the total of the percentages is greater than $100\%$.}

@@

In this chapter, we investigate the energy footprint of Python in its most popular domains of adoption.
We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{hasan_energy_2016} did in~\cite{hasan_energy_2016}.
Then, we measure the energy consumption of several Python implementations to propose a non-intrusive technique to improve energy efficiency.
Therefore, this chapter will address the following research questions:
\begin{compactenum}[\indent\bf RQ\,1:]
    \item \emph{What is the energy footprint of Python when used in data science?}
    \item \emph{Are the Python guidelines energy-efficient by construction?}
    \item \emph{Can we reduce the energy consumption of Python programs without altering the source code?}
\end{compactenum}

@@

Django\footnote{\url{https://www.djangoproject.com/}} and Flask\footnote{\url{https://flask.palletsprojects.com/}} are the most popular frameworks for web development.
According to a Jetbrains poll,\footnote{\url{https://lp.jetbrains.com/python-developers-survey-2021/}} Django is used in 40\% of the cases, while Flask is used in 41\% of the cases.

In contrast to Flask, which is a micro-framework, Django is a high-level web framework that provides a standard method for creating and maintaining complex and scalable database-driven websites quickly and effectively.
Therefore, this study will focus on the latter one.

\subsection{Life-cycle of a Request in Django}
Django is a MVT (\emph{Model-View-Template}) framework, which means that it follows the MVC (\emph{Model-View-Controller}) pattern.
Figure~\ref{fig:django-life-cycle} describes the life-cycle of a request in this framework.
Whenever a request arrives in Django, it is processed by the middleware layers, one at a time.
These middleware layers are in charge of authentication, security, and so on.
Once the request is processed by these layers, it is passed to the URL router, which extracts the URL from the request and tries to match it to the defined URLs.
After getting the matching URL, the corresponding view function is called.
This function is responsible for treating the request, gathering the data, and then generating the response that will be put inside a template to be returned as a response.
As Django adopts a MVT model, it also offers an automatic way to retrieve data from the database to the view using the \emph{Object Relational Mapping} (ORM)~\cite{o2008object}.

@@

First, we start by investigating the energy consumption of the request-response life cycle in Django, to determine which layer consumes the most energy.
To do so, we created a sample Django application that returns the response of the request. We tracked its energy consumption using JouleHunter.\footnote{\url{https://github.com/powerapi-ng/joulehunter}}
JouleHunter is an open-source library that I developed to help practitioners to identify  energy hotspots in their applications using statistical profilers.
In the case of Django, it can be added as an extension with no additional setup or change to the source code.
The energy consumption of the request-response life-cycle in Django is shown in Figure~\ref{fig:django_life_cycle_naive}.
As we can notice, 91.4\% of the total energy consumption is spent to resolve the request by retrieving the data, while only 5.27\% is spent to render the response.

@@

%% REMARQUE the experiment with joulehunter isn't conducted within the same machine that is used for generating the boxplot graphs 
% We have chosen Django as our web framework.
Therefore, we chose to study if the choice of the database and the ORM impacts the energy consumption and the performance of the website.
% \subsection{Experiment \& Results}
To accomplish so, we examined the cost of a single request of the prior website using various ways to extract data from the database.
We considered using two different databases, \textsc{PostgreSQL} and \textsc{SQlite3}, that store the same records, and three different ways to fetch the records:
\begin{enumerate}
    \item \textsf{Vanilla} relies only on the ORM to retrieve the data,
    \item \textsf{Prefetch} queries the data before being requested,
    \item \textsf{Optimized} leverages SQL without passing by the ORM.
\end{enumerate}

%%%%% have to rephrase this 
As one can observe in \Cref{fig:django}, the strategy to query stored data has a huge impact on energy consumption.
As the \textsf{Vanilla} strategy can consume up to $10\times$ more energy than the \textsf{Optimized} one.
Conversely, the choice of the database does not exhibit a key impact on the total energy, despite their different behavior regarding the execution time and the average power.
This can be useful to support developers in choosing which database engine they can adopt, guided by the number of expected requests and the targeted performance.

@@

Another interesting observation is the impact of the interpreter, as \Cref{fig:django} highlights.
For example, using the \textsf{PyPy} interpreter reduces the energy consumption, even when adopting the \textsf{Vanilla} strategy.

Finally, we run the same experiment with the \textsf{Optimized} strategy using JouleHunter.
Figure~\ref{fig:django_profiled_optimized} depicts the resulting energy consumption.
As one can notice, while the rendering method consumed the same amount of energy as in the \textsf{Vanilla} strategy (around 1.3\,kJ), the resolve part dropped by $20\times$.

% \subsection{Conclusion}
We can therefore conclude that the database and ORM selection have a major impact on the website's performance and energy consumption.
And, unlike the rendering portion, there are many options available for the database, the ORM, and the data handling strategy.

\begin{figure}[hbt]
    \centering
    \includegraphics[width=\linewidth]{imgs/django_profiled_optimized}
    \caption{Tree representation of the energy consumption of a single request in Django (naive version)}
    \label{fig:django_profiled_optimized}
\end{figure}

@@

\section{Python Insights}
This section aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
Therefore, we start by extending the work of \citeauthor{hasan_energy_2016} and
\citeauthor{oliveira_recommending_nodate} to the context of Python environments
\cite{hasan_energy_2016,oliveira_recommending_nodate}.

\subsection{experiment and results}
%%%% concurency 
Another field of investigation is the type of data and control structures that might impact the energy consumption.
We iterate over a list using 3 methods.
First, the classical \texttt{for(i in range(len(n)))}.
However, as we can see here, unlike other programming languages, it requires extra operations, such as determining the length of the collection and then using the iterator range.
So, we tried the more adapted version \texttt{for(element in collection)}.
Moreover, in most programming languages. the \texttt{for} loop is translated to a \texttt{while} loop ( transformation from D type to B type -- asm -- ), therefore we wanted to compare this with a \texttt{while} version.
After determining the main ways to iterate over a \texttt{loop}, we run the collection algorithms of different data types to observe if they impact the energy consumption of the code, and we repeated it for the size of the collection.

@@

\subsection{Synthesis}
This study demonstrates that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.

\section{Python \& Multiprocessing}

\subsubsection{Cost of Parallelism in Python}
Additionally, \Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications when introducing parallelism.
Python applications are single-threaded systems constrained by the \emph{global lock system} (GLS).
However, due to the increase of cores/threads per CPU, many Python libraries started to take advantage of this hardware feature by allowing concurrent execution of Python processes.
For example, the multiprocessing library\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}} spawns sub-processes to increase the degree of concurrency of Python applications.
As one can see in \Cref{fig:python_multiprocessing}, the energy consumption is correlated with the number of threads until reaching the limit of physical cores, when concurrent processes start to compete for the CPU.

\Cref{fig:python_multiprocessing} compares the behavior of python programs when we try to introduce parallelism.
As we know python is a single-threaded program thanks to the GLS (\emph{Global Lock System}) however, due to the increase in the number of cores/threads per CPU, many libraries started to take advantage of this hardware feature by simulating multithreading as multiple instances of the processor. % TODO: TO VERIFY
